---
title: "Project--Submission 1"
output: pdf_document
date: "2024-07-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Submission 1 (7/30)

1.Identify one gene, one continuous covariate, and two categorical covariates in the provided dataset.
Note: Gene expression data and metadata are in two separate files and will need to be linked.

``` {r}
#Load packages
library(tidyverse)

#Set the working directory to the path of given dataset.
getwd()

#Using read.csv to read in the gene expression data and metadata
genes_expression<-read.csv(file="QBS103_GSE157103_genes.csv",header=T,stringsAsFactors=F)
metadata<-read.csv(file="QBS103_GSE157103_series_matrix.csv",header=T,stringsAsFactors=F)

#Using 'which' to select the chosen gene and identify it
gene_AAK1<-genes_expression[which(genes_expression[,1]=='AAK1'),]

#use pipe and merge to convert the gene expression to required format and link two dataframes
gene_AAK1<-gene_AAK1 %>%
  gather(key=participant_id,value=expression)
linked_data<-merge(metadata,gene_AAK1)
#head(linked_data)


```

2.Generate the following three plots using ggplot2 for your covariates of choice:
  2.1 Histogram for gene expression (5 pts)
```{r}
ggplot(linked_data,aes(x=as.numeric(expression)))+
  geom_histogram(fill='lightblue')+
  labs(title="Gene Expression of AAK1",
       x="AAK1",
       y="Count")
```
  
  2.2 Scatterplot for gene expression and continuous covariate (5 pts)
```{r}
linked_data$expression<-as.numeric(linked_data$expression)
ggplot(linked_data,aes(x=ddimer.mg.l_feu.,y=expression))+
  geom_point(color='darkblue')+
  labs(title='Scatterplot for AAK1 expression and ddimer',x='ddimer',y='AAK1 expression')
  
```
  
  2.3 Boxplot of gene expression separated by both categorical covariates (5 pts)
```{r}
ggplot(linked_data,aes(x=sex,y=expression,fill = icu_status))+
  geom_boxplot()+
  labs(title="Boxplot of AAK1 Expression separated by Sex and ICU Status",
       x="Sex",
       y="AAK1 Expression")
```

