---
title: "240807 Submission 2"
output: pdf_document
date: "2024-08-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Submission 2
1.Build a function to create the plots you made for Presentation 1, incorporating feedback you received from your first presentation on improving plot design. Your functions should take the following input: (1) the name of the data frame, (2) a list of 1 or more gene names, (3) 1 continuous covariate, and (4) two categorical covariates (10 pts)
```{r}
library(tidyverse)
library(dplyr)
```

```{r}
makePlot<-function(dataSet,geneName,Cont,Cat1,Cat2){
  metadata<-read.csv(file="QBS103_GSE157103_series_matrix.csv",header=T,stringsAsFactors=F)
  
  #initialize an empty list to store the plots
  plots<-list()
  
  for (gene in geneName){

    #using 'which' to select the chosen gene and identify it
    gene_selected<-dataSet[which(dataSet[,1]==gene),]
    
    #use pipe and merge to convert the gene expression to required format and link two dataframes
    gene_selected<-gene_selected %>%
      gather(key=participant_id,value=expression)
    linked_data<-merge(metadata,gene_selected)
    
    #change the gene expression of the linked data frame into numeric type
    linked_data$expression<-as.numeric(linked_data$expression)
    linked_data$ddimer.mg.l_feu.<-as.numeric(linked_data$ddimer.mg.l_feu.)
    
    #generate the histogram plot
    p1<-ggplot(linked_data,aes(expression))+
      geom_histogram(fill='lightblue',color='#694F8E')+
      scale_x_continuous(breaks = seq(0,100,by=1)) +
      labs(title=paste0("Gene Expression of ",gene),x=gene,y='Count')+
      theme_classic()+
      theme(
        plot.title=element_text(hjust=0.5,size=20,face='bold'),
        axis.title.x = element_text(size = 12, face = "bold"),  
        axis.title.y = element_text(size = 12, face = "bold")
      )
      
    # Set x-axis to display specified ticks
    breaks <- seq(0, 200, by = 5)  # Display every 5 ticks
    
    #generate the scatterplot and do customization
    p2<-ggplot(linked_data,aes(y=expression,x=ddimer.mg.l_feu.))+
      geom_point(color='darkblue')+
      scale_x_continuous(breaks = breaks) +
      labs(title=paste0('Scatterplot for ',gene,' and ddimer'),y=gene,x='ddimer(mg/L_feu)')+
      theme_classic()+
      theme(
        plot.title = element_text(hjust = 0.5, size = 20, face = "bold"),
        axis.title.x = element_text(size = 12, face = "bold"),  
        axis.title.y = element_text(size = 12, face = "bold"),  
        #turn a angle to avoid overlapping
        axis.text.x = element_text(size = 8, angle = 45, hjust = 1), 
        axis.text.y = element_text(size = 10)
        )

    #generate the boxplot 
    p3<-ggplot(linked_data,aes(x=sex,y=expression,fill=icu_status))+
      geom_boxplot(outlier.shape = NA)+
      labs(title=paste0("Boxplot of ",gene," Expression separated by ",Cat1," and ",Cat2),
       x=Cat1,
       y=gene)+
      stat_summary(fun = mean, geom = "point", shape = 20, size = 3, color = "#A94438", 
                   position = position_dodge(width = 0.75))+
      scale_fill_manual(values=c(" yes"='#EED3D9'," no"='#B5C0D0'))+
      theme_classic()+
      theme(
        plot.title = element_text(hjust = 0.5, size = 15, face = "bold"),
        axis.title.x = element_text(size = 12, face = "bold"),  
        axis.title.y = element_text(size = 12, face = "bold"),  
        axis.text.x = element_text(size = 10), 
        axis.text.y = element_text(size = 10),
        legend.title = element_text(size = 12),  
        legend.text = element_text(size = 10),
        legend.position ="top"
      )
    
    #add the three plots into the list according to its input gene name
    plots[[gene]] <- list(histogram = p1, scatter = p2, boxplot = p3)
    }
    
    return(plots)
}

```


2.Select 2 additional genes (for a total of 3 genes) to look at and implement a loop to generate your figures using the function you created (10 pts)
```{r}
setwd("E:/55555 Dartmouth QBS/QBS 103 Foundation of DS(R)/Project")
genes_expression<-read.csv(file="QBS103_GSE157103_genes.csv",header=T,stringsAsFactors=F)

# Specify genes of interest
genes <- c('AAK1', 'A1BG', 'AAAS') 

# Generate plots
plots <- makePlot(dataSet=genes_expression, geneName=genes, Cont='ddimer.mg.l_feu.', 
                  Cat1='sex',
                  Cat2='icu_status')

# Display the plots
for (gene in names(plots)) {
  print(plots[[gene]]$histogram)
  print(plots[[gene]]$scatter)
  print(plots[[gene]]$boxplot)
}
```
